---
title: "roptuna: A Practitioner's Guide to Hyperparameter Optimization"
output:
  rmarkdown::html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: pygments
    css: roptuna-vignette.css
vignette: >
  %\VignetteIndexEntry{roptuna: A Practitioner's Guide to Hyperparameter Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4,
  fig.align  = "center"
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(roptuna)
library(ggplot2)
set.seed(42)
```

## Introduction

Every supervised learning model carries hyperparameters — settings fixed before training begins that the learning algorithm itself cannot determine. Learning rate, tree depth, regularisation strength, dropout rate, and kernel bandwidth all fall into this category. Choosing them well is not optional: a support vector machine with the wrong kernel width will perform far below its potential on the same data where a carefully tuned version excels, and the gap widens as the model and dataset grow.

Manual tuning does not scale. A modern gradient boosting model exposes a dozen or more hyperparameters; their joint search space is too irregular for grid search or researcher intuition to navigate efficiently. **Bayesian optimisation** addresses this by treating hyperparameter selection as a sequential decision problem: after each trial the algorithm updates a probabilistic model of the objective surface and uses it to guide the next sample, concentrating effort in regions likely to improve on the current best.

`roptuna` is an R implementation of the [Optuna](https://optuna.org/) framework (Akiba et al. 2019). It brings Optuna's *define-by-run* API to native R: instead of declaring the search space as a configuration object before optimisation begins, you call `trial$suggest_*()` inside your objective function, and the sampler adapts as evidence accumulates. This design accommodates **conditional search spaces** — where which parameters are active depends on the value of a prior parameter — naturally and without extra machinery.

---

## The R HPO Ecosystem

Several packages address hyperparameter optimisation in R, each with different designs and tradeoffs.

| Package | Method | Interface | Pruning | Notes |
|---|---|---|---|---|
| `roptuna` | TPE, Random, Grid | Define-by-run | Yes | This package |
| `mlr3tuning` | Many (via learner) | mlr3 ecosystem | Via callbacks | Framework integration |
| `tune` | Many (via engine) | tidymodels ecosystem | No | Framework integration |
| `ParBayesianOptimization` | Gaussian process BO | Formula-based | No | GP-based; separate search space |
| `rBayesianOptimization` | Gaussian process BO | Formula-based | No | Simple single-file API |
| `irace` | Iterated racing | Config-based | No | Algorithm configuration focus |

`roptuna` is a **standalone sampler library**: it does not require mlr3 or tidymodels, but provides adapters for both so it can serve as the optimisation engine inside either framework. For teams already invested in those ecosystems the adapters allow `roptuna`'s samplers to be dropped in without changing workflow, resampling, or preprocessing code.

---

## How TPE Works

The default sampler is the **Tree-structured Parzen Estimator** (TPE), introduced by Bergstra et al. (2011). Understanding what it does internally helps you use it well and diagnose cases where it under-performs.

### The core idea

After a configurable number of startup trials (random exploration), TPE divides past trials into two groups based on their objective value:

- **Good trials** (*l* group): the best γ fraction of completed trials (default γ = 0.25)
- **Bad trials** (*g* group): the remaining 1 − γ fraction

For each parameter, TPE fits a **kernel density estimate** over the parameter values that appeared in each group: $\ell(x)$ over good trials and $g(x)$ over bad trials. It then generates a set of candidate parameter values and selects the candidate where the **expected improvement score** $\ell(x) / g(x)$ is highest.

The intuition is direct: a region where $\ell(x)$ is high and $g(x)$ is low is a region where good trials concentrate and bad trials do not — exactly where the next evaluation should land.

### Visualising the mechanism

The scatter plot below shows 20 past trials of a 1-D quadratic objective. Points are coloured by whether they belong to the good (bottom 25%) or bad (top 75%) group.

```{r tpe-scatter, fig.height = 4.5, message = FALSE, warning = FALSE}
set.seed(7)
x_obs <- runif(20, -5, 5)
y_obs <- x_obs^2 + rnorm(20, sd = 0.5)

gamma     <- 0.25
threshold <- quantile(y_obs, gamma)
group     <- ifelse(y_obs <= threshold, "Good (bottom 25%)", "Bad (top 75%)")

ggplot(data.frame(x = x_obs, y = y_obs, group = group),
       aes(x = x, y = y, colour = group, shape = group)) +
  geom_point(size = 3.2, alpha = 0.85) +
  geom_hline(yintercept = threshold, linetype = "dashed", colour = "grey45") +
  annotate("text", x = 4.0, y = threshold + 1.4,
           label = sprintf("gamma threshold = %.1f", threshold),
           colour = "grey40", size = 3.2) +
  scale_colour_manual(values = c("Good (bottom 25%)" = "#2171b5",
                                 "Bad (top 75%)"     = "#ef6548")) +
  scale_shape_manual(values  = c("Good (bottom 25%)" = 16,
                                 "Bad (top 75%)"     = 1)) +
  labs(
    title    = "TPE: classifying past trials into 'good' and 'bad'",
    subtitle = "Good trials (bottom 25%) drive the density estimate l(x); bad trials drive g(x)",
    x        = "Parameter value (x)", y = "Objective value f(x)",
    colour   = NULL, shape = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(colour = "grey40"))
```

Given these two groups, TPE fits a kernel density estimate for each and evaluates the acquisition function $\ell(x) / g(x)$ across the domain. The vertical dotted line marks where this ratio peaks — the suggested location for the next trial.

```{r tpe-density, fig.height = 4, message = FALSE, warning = FALSE}
x_good <- x_obs[y_obs <= threshold]
x_bad  <- x_obs[y_obs >  threshold]
x_grid <- seq(-5, 5, length.out = 300)
bw     <- 1.2

l_vals  <- sapply(x_grid, function(xi) mean(dnorm(xi, x_good, bw)))
g_vals  <- sapply(x_grid, function(xi) mean(dnorm(xi, x_bad,  bw)))
ei_vals <- l_vals / (g_vals + 1e-9)

l_norm  <- l_vals  / max(l_vals)
g_norm  <- g_vals  / max(g_vals)
ei_norm <- (ei_vals - min(ei_vals)) / diff(range(ei_vals))
next_x  <- x_grid[which.max(ei_vals)]

df_dens <- rbind(
  data.frame(x = x_grid, y = l_norm,  curve = "l(x) — good trial density"),
  data.frame(x = x_grid, y = g_norm,  curve = "g(x) — bad trial density"),
  data.frame(x = x_grid, y = ei_norm, curve = "EI = l(x)/g(x) — acquisition")
)

ggplot(df_dens, aes(x = x, y = y, colour = curve)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = next_x, linetype = "dotted",
             linewidth = 1.1, colour = "grey20") +
  annotate("label", x = next_x + 0.3, y = 0.93,
           label = sprintf("Next sample\nx = %.2f", next_x),
           size = 3.1, fill = "white", label.size = 0.3, colour = "grey20") +
  scale_colour_manual(values = c(
    "l(x) — good trial density"     = "#2171b5",
    "g(x) — bad trial density"      = "#ef6548",
    "EI = l(x)/g(x) — acquisition"  = "#238b45"
  )) +
  labs(
    title    = "TPE: density estimates and acquisition function",
    subtitle = "Next trial is placed at the argmax of l(x)/g(x)",
    x        = "Parameter value (x)", y = "Normalised value",
    colour   = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(colour = "grey40"))
```

---

## Getting Started

### Installation

```r
# From CRAN:
install.packages("roptuna")

# Development version from GitHub:
remotes::install_github("kvenkita/roptuna")
```

### A minimal example

```{r quick-start, message = FALSE}
objective <- function(trial) {
  x <- trial$suggest_float("x", -5, 5)
  y <- trial$suggest_int("y", 1L, 5L)
  x^2 + y
}

study <- create_study("minimize", sampler = tpe_sampler(seed = 42))
study$optimize(objective, n_trials = 50)

cat("Best value:", study$best_value, "\n")
cat("Best params:\n"); str(study$best_params)
```

Five functions are all you need for the basic workflow: `create_study()`, `study$optimize()`, `trial$suggest_float()`, `trial$suggest_int()`, and `trial$suggest_categorical()`.

---

## The Study–Trial API

### `create_study()`

```r
study <- create_study(
  direction      = "minimize",    # or "maximize"
  sampler        = tpe_sampler(), # sampling strategy; random_sampler() by default
  pruner         = NULL,          # early stopping (optional)
  storage        = NULL,          # InMemoryStorage by default
  study_name     = "my-study",    # auto-generated if NULL
  load_if_exists = FALSE          # for SQLite: resume an existing study
)
```

### Study active bindings

Once a study has been optimised, the following properties expose results:

```{r study-properties}
study_props <- create_study("minimize", sampler = tpe_sampler(seed = 99))
study_props$optimize(function(trial) {
  x <- trial$suggest_float("x", -10, 10)
  x^2
}, n_trials = 30)

cat("Best value:   ", study_props$best_value, "\n")
cat("Best x:       ", study_props$best_params$x, "\n")
cat("Total trials: ", length(study_props$trials), "\n")
cat("State table:\n"); print(table(sapply(study_props$trials, `[[`, "state")))
```

### `trial$suggest_*()` methods

| Method | Description | Distribution |
|---|---|---|
| `trial$suggest_float(name, low, high)` | Continuous uniform | *U*(low, high) |
| `trial$suggest_float(name, low, high, log = TRUE)` | Log-uniform | *LogU*(low, high) |
| `trial$suggest_int(name, low, high)` | Discrete uniform | *DiscreteU*{low, …, high} |
| `trial$suggest_categorical(name, choices)` | Nominal choice | Empirical from choices |

**Parameters are idempotent within a trial.** Calling `suggest_float("lr", 0, 1)` twice in the same trial always returns the same value. This is useful when a shared sub-function appears multiple times in a complex objective.

**Log-scale parameters** should be used whenever a quantity spans more than one order of magnitude. Learning rates (*10⁻⁵* to *10⁻¹*), regularisation penalties, and kernel bandwidths are the most common examples. On a linear scale, TPE's Parzen estimate concentrates probability mass at the high end of such a range; log scale distributes exploration evenly across magnitudes.

```{r suggest-demo}
study_demo <- create_study("minimize", sampler = tpe_sampler(seed = 7))
study_demo$optimize(function(trial) {
  lr     <- trial$suggest_float("lr",     1e-5, 1e-1, log = TRUE)
  depth  <- trial$suggest_int("depth",    2L, 8L)
  method <- trial$suggest_categorical("method", c("adam", "sgd", "rmsprop"))
  penalty <- c(adam = 0, sgd = 0.3, rmsprop = 0.6)[[method]]
  (log10(1 / lr) - 3)^2 * 0.1 + 0.05 * depth + penalty
}, n_trials = 40)

cat("Best method:", study_demo$best_params$method, "\n")
cat("Best lr:    ", formatC(study_demo$best_params$lr, format = "e", digits = 2), "\n")
cat("Best depth: ", study_demo$best_params$depth, "\n")
```

---

## Samplers

### Available samplers

| Sampler | Function | When to use |
|---|---|---|
| **TPE** | `tpe_sampler(seed, n_startup_trials, gamma, n_ei_candidates)` | General purpose; default choice |
| **Random** | `random_sampler(seed)` | Baselines; parallel embarrassingly-random search |
| **Grid** | `grid_sampler(search_space)` | Exhaustive small grids; all combinations required |

### TPE vs Random: convergence comparison

The practical advantage of TPE over random search is most visible when good solutions occupy a small fraction of the search space. The 2-D bowl below has a clear basin near (0, 0) that TPE locates efficiently while random search takes longer.

```{r sampler-comparison, fig.height = 4.5}
study_tpe <- create_study("minimize", sampler = tpe_sampler(seed = 123))
study_tpe$optimize(function(trial) {
  x <- trial$suggest_float("x", -10, 10)
  y <- trial$suggest_float("y", -10, 10)
  x^2 + y^2 + 0.5 * x * y    # bowl with a ridge
}, n_trials = 60)

study_rnd <- create_study("minimize", sampler = random_sampler(seed = 456))
study_rnd$optimize(function(trial) {
  x <- trial$suggest_float("x", -10, 10)
  y <- trial$suggest_float("y", -10, 10)
  x^2 + y^2 + 0.5 * x * y
}, n_trials = 60)

extract_history <- function(study, label) {
  vals <- sapply(study$trials, `[[`, "value")
  data.frame(trial = seq_along(vals), value = vals,
             best = cummin(vals), sampler = label)
}

hist_df <- rbind(
  extract_history(study_tpe, "TPE"),
  extract_history(study_rnd, "Random")
)

ggplot(hist_df, aes(x = trial)) +
  geom_point(aes(y = value, colour = sampler), alpha = 0.22, size = 1.3) +
  geom_line(aes(y = best, colour = sampler), linewidth = 1.2) +
  scale_colour_manual(values = c("TPE" = "#2171b5", "Random" = "#ef6548")) +
  labs(
    title    = "TPE vs Random: cumulative best on a 2-D bowl",
    subtitle = "Lines show the running best; points show individual trial objective values",
    x = "Trial number", y = "Objective value", colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(colour = "grey40"))
```

### Grid sampler

The grid sampler exhaustively evaluates every combination of a declared search space. It is appropriate when the space is small and complete coverage matters more than sample efficiency — for example, when ablating a small set of architectural choices.

```{r grid-sampler}
study_grid <- create_study("minimize",
  sampler = grid_sampler(list(
    n_estimators = c(50L, 100L, 200L),
    max_depth    = c(3L,  5L,   7L)
  ))
)
study_grid$optimize(function(trial) {
  n <- trial$suggest_categorical("n_estimators", c(50L, 100L, 200L))
  d <- trial$suggest_categorical("max_depth",    c(3L,  5L,  7L))
  0.35 - 0.001 * as.numeric(n) + 0.01 * abs(as.numeric(d) - 4)
}, n_trials = 9)  # 3 × 3 = 9 combinations

grid_df <- do.call(rbind, lapply(study_grid$trials, function(t)
  data.frame(n_estimators = t$params$n_estimators,
             max_depth    = t$params$max_depth,
             value        = round(t$value, 4))))
print(grid_df[order(grid_df$value), ])
```

---

## Pruning

Pruning terminates unpromising trials early, before they consume their full evaluation budget. A trial whose intermediate performance is clearly worse than existing trials can be abandoned after a fraction of its epochs, freeing compute for more promising candidates.

### How to add pruning

Your objective function reports intermediate values with `trial$report()`, checks `trial$should_prune()`, and signals early termination with `stop_prune()`.

```{r pruning-demo}
study_prune <- create_study(
  "minimize",
  sampler = tpe_sampler(seed = 42),
  pruner  = median_pruner(n_startup_trials = 5L, n_warmup_steps = 3L)
)

study_prune$optimize(function(trial) {
  lr <- trial$suggest_float("lr", 1e-4, 1e-1, log = TRUE)
  for (epoch in seq_len(20)) {
    loss <- exp(-lr * epoch) + 0.1 + rnorm(1, 0, 0.03)
    trial$report(loss, step = as.integer(epoch))
    if (trial$should_prune()) stop_prune()
  }
  loss
}, n_trials = 40)

cat("Trial state distribution:\n")
print(table(sapply(study_prune$trials, `[[`, "state")))
```

### Visualising the pruning effect

Training curves for all trials reveal the pruner's behaviour: trials with high loss early (low learning rate, slow convergence) are cut short, while trials that converge quickly (high learning rate) run to completion.

```{r pruning-viz, fig.height = 5}
traces <- Filter(Negate(is.null), lapply(study_prune$trials, function(t) {
  iv <- t$intermediate_values
  if (length(iv) == 0) return(NULL)
  data.frame(
    trial  = t$number,
    epoch  = as.integer(names(iv)),
    loss   = unlist(iv),
    state  = t$state,
    stringsAsFactors = FALSE
  )
}))
trace_df <- do.call(rbind, traces)
trace_df$label <- ifelse(trace_df$state == "pruned", "Pruned", "Completed")

ggplot(trace_df, aes(x = epoch, y = loss,
                     group = trial, colour = label)) +
  geom_line(alpha = 0.55, linewidth = 0.65) +
  scale_colour_manual(values = c("Pruned" = "#ef6548", "Completed" = "#2171b5")) +
  labs(
    title    = "Trial training curves with MedianPruner",
    subtitle = "Pruned trials (red) are stopped when their loss exceeds the running median",
    x        = "Training epoch", y = "Loss", colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        plot.title      = element_text(face = "bold"),
        plot.subtitle   = element_text(colour = "grey40"))
```

### MedianPruner vs SuccessiveHalvingPruner

| Pruner | Function | Mechanism |
|---|---|---|
| **Median** | `median_pruner(n_startup_trials, n_warmup_steps, interval_steps)` | Prune if loss > median of completed trials at the same step |
| **Successive Halving (ASHA)** | `successive_halving_pruner(min_resource, reduction_factor)` | Promote only the top 1/η fraction to the next resource rung |

The MedianPruner is the safer default: it requires relatively few completed trials before activating and does not depend on a predetermined total budget. The SuccessiveHalvingPruner is more aggressive and works best when individual trials are long and you have a large overall evaluation budget.

---

## Visualisation

`roptuna` implements `autoplot.Study` with three plot types. All three accept the `Study` object returned by `create_study()`.

### Setup: a multi-parameter study

```{r viz-study}
study_viz <- create_study("minimize", sampler = tpe_sampler(seed = 42))
study_viz$optimize(function(trial) {
  lr      <- trial$suggest_float("lr",      1e-4, 1.0, log = TRUE)
  dropout <- trial$suggest_float("dropout", 0.0,  0.8)
  depth   <- trial$suggest_int("depth",     1L,   5L)
  (1 - lr)^2 + 100 * (dropout - lr^2)^2 + 0.5 * depth
}, n_trials = 80)
```

### Optimization history

```{r plot-history}
autoplot(study_viz, type = "history") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

The history plot shows every trial's objective value (grey points) and the running best (blue line). The characteristic signature of TPE is a steep early drop — rapid improvement once the sampler has enough data to model the surface — followed by a long plateau as the basin has been well-explored.

### Parallel coordinates

```{r plot-parallel}
autoplot(study_viz, type = "parallel_coordinate") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

The parallel coordinates plot normalises all hyperparameter values to [0, 1] and draws one line per trial. Lines are coloured by objective value (darker = lower for minimisation). Regions where many good lines converge reveal the most promising area of the joint search space and surface interactions between parameters that a univariate analysis would miss.

### Parameter importance

```{r plot-importance}
autoplot(study_viz, type = "param_importance") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

Parameter importance is estimated by fitting a shallow decision tree (`rpart`) to the completed trials and reporting the variable importance scores. Parameters with near-zero importance can often be fixed at their default values in subsequent rounds, effectively reducing the search space and making each remaining trial more informative.

---

## Persistent Studies with SQLite

By default, `roptuna` stores trial results in memory and they are lost when the R session ends. SQLite persistence provides a durable, queryable record that survives session restarts and can be shared across machines.

```{r sqlite-demo, message = FALSE}
tmp_db <- tempfile(fileext = ".sqlite")

# First session: run 20 trials
study_s1 <- create_study(
  "minimize",
  sampler    = tpe_sampler(seed = 11),
  storage    = sqlite_storage(tmp_db),
  study_name = "persistent-demo"
)
study_s1$optimize(function(trial) trial$suggest_float("x", -5, 5)^2,
                  n_trials = 20)
cat("Session 1 best:", round(study_s1$best_value, 5), "\n")

# Second session: reload and add 20 more trials
study_s2 <- create_study(
  "minimize",
  sampler        = tpe_sampler(seed = 22),
  storage        = sqlite_storage(tmp_db),
  study_name     = "persistent-demo",
  load_if_exists = TRUE
)
study_s2$optimize(function(trial) trial$suggest_float("x", -5, 5)^2,
                  n_trials = 20)
cat("Session 2 total trials:", length(study_s2$trials), "\n")
cat("Overall best:          ", round(study_s2$best_value, 6), "\n")

unlink(tmp_db)
```

The `load_if_exists = TRUE` flag finds an existing study by name in the database rather than creating a fresh one. All prior trials are immediately available through `study$trials`, and the new sampler uses them to inform its subsequent suggestions — the optimisation continues where the previous session left off.

---

## A Realistic Example: Tuning a Decision Tree on Iris

This section demonstrates a complete HPO workflow on a real dataset: a three-hyperparameter search including a log-scale parameter, 5-fold cross-validation as the objective, and all three plot types applied to the results.

```{r rpart-setup, message = FALSE}
set.seed(42)

iris_objective <- function(trial) {
  minsplit <- trial$suggest_int("minsplit", 2L, 40L)
  maxdepth <- trial$suggest_int("maxdepth", 1L, 12L)
  cp       <- trial$suggest_float("cp", 0.0001, 0.5, log = TRUE)

  folds <- sample(rep(1:5, length.out = nrow(iris)))
  acc <- vapply(1:5, function(k) {
    tr  <- iris[folds != k, ]
    te  <- iris[folds == k, ]
    fit <- rpart::rpart(
      Species ~ ., data = tr, method = "class",
      control = rpart::rpart.control(
        minsplit = minsplit, maxdepth = maxdepth, cp = cp
      )
    )
    mean(predict(fit, te, type = "class") == te$Species)
  }, numeric(1))
  1 - mean(acc)
}

study_iris <- create_study("minimize", sampler = tpe_sampler(seed = 42))
study_iris$optimize(iris_objective, n_trials = 60)

cat("Best CV error rate:", round(study_iris$best_value, 4), "\n")
cat("Best params:\n"); str(study_iris$best_params)
```

```{r rpart-history, fig.height = 3.5}
autoplot(study_iris, type = "history") +
  labs(subtitle = "iris CART: 5-fold CV error over 60 trials") +
  theme_minimal(base_size = 11) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"))
```

```{r rpart-parallel, fig.height = 3.5}
autoplot(study_iris, type = "parallel_coordinate") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

```{r rpart-importance, fig.height = 3}
autoplot(study_iris, type = "param_importance") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

The importance plot reveals which of the three hyperparameters most influences cross-validated accuracy. On iris, `cp` (the complexity parameter that sets the minimum improvement required for a split) typically dominates — the tree structure is more sensitive to over-splitting than to maximum depth alone, because iris is nearly linearly separable and shallow trees suffice.

---

## Framework Adapters

`roptuna` provides adapters for the two major R modelling frameworks. These are thin wrappers that plug `roptuna`'s samplers into the framework's existing workflow, resampling, and metric infrastructure.

### tidymodels

```{r tidymodels-adapter, eval = FALSE}
library(tune); library(parsnip); library(rsample); library(workflows)

dt_spec <- decision_tree(
  min_n = tune(), tree_depth = tune(), cost_complexity = tune()
) |>
  set_engine("rpart") |>
  set_mode("classification")

wf  <- workflow() |> add_model(dt_spec) |> add_formula(Species ~ .)
cv  <- vfold_cv(iris, v = 5)

res <- tune_optuna(wf, cv,
  direction = "minimize",
  metric    = "accuracy",
  n_trials  = 40,
  sampler   = tpe_sampler(seed = 1L)
)
tune::show_best(res)
```

### mlr3

```{r mlr3-adapter, eval = FALSE}
library(mlr3); library(mlr3tuning); library(paradox)

task    <- tsk("iris")
learner <- lrn("classif.rpart",
               minsplit = to_tune(2L, 40L),
               maxdepth = to_tune(1L, 12L))
tuner   <- TunerOptuna$new(direction = "minimize",
                            sampler   = tpe_sampler(seed = 1L))
ti <- tune(
  tuner      = tuner,
  task       = task,
  learner    = learner,
  resampling = rsmp("cv", folds = 5L),
  measures   = msr("classif.ce"),
  term_evals = 40L
)
ti$result
```

---

## Practical Guidance

### How many trials?

Allow roughly **10–20 trials per active hyperparameter** for TPE to move past its startup phase and build a meaningful surface model. For three hyperparameters, 50–100 trials is a reasonable starting budget. If the objective is cheap (under a second per trial), run 200 or more; the improvement in sample efficiency compounds. For expensive objectives (minutes per trial), consider starting with random search to build a baseline, then switching to TPE.

### Choosing n_startup_trials

The `n_startup_trials` argument (default 10) controls how many random trials are run before TPE activates. Increasing it reduces the risk of overfitting to an unlucky initial sample; decreasing it starts Bayesian exploitation earlier but on thinner evidence. For fewer than 3 active parameters, `n_startup_trials = 5` is usually sufficient. For 10 or more parameters, consider `n_startup_trials = 20`.

### Conditional search spaces

`roptuna`'s define-by-run design handles conditional parameters naturally — only call `suggest_*()` for parameters that are active given earlier choices.

```{r conditional}
study_cond <- create_study("minimize", sampler = tpe_sampler(seed = 42))
study_cond$optimize(function(trial) {
  model <- trial$suggest_categorical("model", c("linear", "tree"))
  if (model == "linear") {
    alpha <- trial$suggest_float("alpha", 0.0001, 10, log = TRUE)
    0.5 + log10(alpha + 1) * 0.08
  } else {
    depth <- trial$suggest_int("depth", 1L, 8L)
    0.3 + 0.05 * depth
  }
}, n_trials = 40)

cat("Best model:", study_cond$best_params$model, "\n")
if (study_cond$best_params$model == "tree") {
  cat("Best depth:", study_cond$best_params$depth, "\n")
} else {
  cat("Best alpha:", round(study_cond$best_params$alpha, 4), "\n")
}
```

Parameters that were never suggested in the best trial simply do not appear in `best_params`. This is the expected behaviour for conditional spaces and requires no special handling.

### Handling objective failures

If the objective occasionally fails — NaN loss, numerical overflow, an out-of-memory error on an extreme hyperparameter combination — use the `catch` argument to absorb specific error classes rather than aborting the study:

```{r catch-errors}
study_robust <- create_study("minimize")
study_robust$optimize(
  function(trial) {
    x <- trial$suggest_float("x", -5, 5)
    if (x > 4.2) stop("simulated domain error")
    x^2
  },
  n_trials = 25,
  catch    = "simpleError"
)
cat("State distribution:\n")
print(table(sapply(study_robust$trials, `[[`, "state")))
```

Failed trials are stored with state `"failed"` and excluded from TPE's model. The study continues without interruption, and the failed trial count gives you a signal that the extreme region of the search space is problematic.

### Using a timeout

For production HPO where wall-clock time matters more than a fixed trial count, use `timeout` (seconds) instead of or in addition to `n_trials`:

```r
study$optimize(objective, n_trials = 1000L, timeout = 3600)  # stop after 1 hour
```

The loop stops at the end of the first trial that completes after the timeout — no trial is interrupted mid-evaluation.

---

## Citation

If you use `roptuna` in published research, please cite both the package and the Optuna paper:

```bibtex
@software{Venkitasubramanian2026roptuna,
  author = {Venkitasubramanian, Kailas},
  title  = {{roptuna}: Hyperparameter Optimization Using the {Optuna} Framework},
  year   = {2026},
  url    = {https://github.com/kvenkita/roptuna},
  note   = {R package version 0.1.0}
}

@inproceedings{Akiba2019optuna,
  author    = {Akiba, Takuya and Sano, Shotaro and Yanase, Toshihiko and
               Ohta, Takeru and Koyama, Masanori},
  title     = {Optuna: A Next-generation Hyperparameter Optimization Framework},
  booktitle = {Proceedings of the 25th {ACM} {SIGKDD} International Conference
               on Knowledge Discovery and Data Mining},
  year      = {2019},
  doi       = {10.1145/3292500.3330701}
}
```

```r
citation("roptuna")
```

**Contact and feedback:** Bug reports and feature requests are welcome at
[kailasv@gmail.com](mailto:kailasv@gmail.com) or via the
[GitHub issue tracker](https://github.com/kvenkita/roptuna/issues).

**Copyright © 2026 Kailas Venkitasubramanian.** Released under the MIT License.

---

## References

Akiba, T., Sano, S., Yanase, T., Ohta, T., and Koyama, M. (2019). Optuna: A next-generation hyperparameter optimization framework. *Proceedings of the 25th ACM SIGKDD International Conference on Knowledge Discovery and Data Mining*, 2623–2631. DOI:10.1145/3292500.3330701.

Bergstra, J., Bardenet, R., Bengio, Y., and Kégl, B. (2011). Algorithms for hyper-parameter optimization. *Advances in Neural Information Processing Systems* 24 (NIPS 2011). https://proceedings.neurips.cc/paper_files/paper/2011/hash/86e8f7ab32cfd12577bc2619bc635690-Abstract.html.

Falkner, S., Klein, A., and Hutter, F. (2018). BOHB: Robust and efficient hyperparameter optimization at scale. *Proceedings of the 35th International Conference on Machine Learning (ICML 2018)*, PMLR 80:1437–1446.

Li, L., Jamieson, K., DeSalvo, G., Rostamizadeh, A., and Talwalkar, A. (2017). Hyperband: A novel bandit-based approach to hyperparameter optimization. *Journal of Machine Learning Research*, 18(1), 6765–6816.
